---
title: "Reproducible Research Project 2"
author: "Nick Daughenbaugh"
date: "3/5/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1: Goal
-----------

The goal of the assignment is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

The following analysis investigates which types of severe weather events are most harmful on:

1.  Health (injuries and fatalities)
2.  Property and crops (economic consequences)

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

2: Data Processing
------------------

### 2.0: Setup

Load the required packages, define useful functions, and set the working directory for this project.

```{r pressure, echo=FALSE}
#Load packages
library(data.table)
library(ggplot2)

#Define a 'not in' function
'%ni%' <- function(x,y)!('%in%'(x,y))

#Define a function 
milbil <- function(x){
  if (x >= 1e9){ paste0("$", round(x/1e9,0), "B")}
  else if (x >= 1e6){ paste0("$", round(x/1e6,0), "M")}
}

#Set working directory
setwd(paste0(getwd(),"/Johns Hopkins University/Reproducible Research/Course Project 1"))
```

### 2.1: Load Data

Download the data, if not already present, and load the data.  Remove columnsthat aren't useful.

```{r pressure, echo=FALSE}
#Download .csv.bz2
if (!file.exists("StormData.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
    destfile = file.path(getwd(), "StormData.csv.bz2"))
}
StormData <- fread("StormData.csv.bz2")
StormData <- StormData[, c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG", "CROPDMGEXP")]
```

### 2.2: Determine Total Economic Damage

Replace, "K", "M", "B" exponents with numeric values; determine total economic damage, Remove columnsthat aren't useful.

```{r pressure, echo=FALSE}
#Property Damage
StormData[(PROPDMGEXP == "K" | PROPDMGEXP == "k"), "PROPDMGEXP"] <- 3
StormData[(PROPDMGEXP == "M" | PROPDMGEXP == "m"), "PROPDMGEXP"] <- 6
StormData[(PROPDMGEXP == "B" | PROPDMGEXP == "b"), "PROPDMGEXP"] <- 9
StormData[(PROPDMGEXP == "H" | PROPDMGEXP == "h"), "PROPDMGEXP"] <- 2
StormData[PROPDMGEXP %ni% 1:9, "PROPDMGEXP"] <- 0
StormData[,Property_Damage := as.numeric(PROPDMG) * 10^(as.numeric(PROPDMGEXP))]

#Storm Damage Exponent
StormData[(CROPDMGEXP == "K" | CROPDMGEXP == "k"), "CROPDMGEXP"] <- 3
StormData[(CROPDMGEXP == "M" | CROPDMGEXP == "m"), "CROPDMGEXP"] <- 6
StormData[(CROPDMGEXP == "B" | CROPDMGEXP == "b"), "CROPDMGEXP"] <- 9
StormData[(CROPDMGEXP == "H" | CROPDMGEXP == "h"), "CROPDMGEXP"] <- 2
StormData[CROPDMGEXP %ni% 1:9, "CROPDMGEXP"] <- 0
StormData[,Crop_Damage := as.numeric(CROPDMG) * 10^(as.numeric(CROPDMGEXP))]

#Economic_Damage
StormData[,Economic_Damage := Property_Damage + Crop_Damage]

#Remove Feilds
StormData[,c("CROPDMG", "CROPDMGEXP", "PROPDMG", "PROPDMGEXP") := NULL]
```

### 2.4: Summerize by Event

```{r pressure, echo=FALSE}
#Summerize Economic Damage
Damage <- StormData[,sum(Economic_Damage, na.rm = TRUE), EVTYPE][,Event := EVTYPE][,Economic_Damage := V1][,c("EVTYPE","V1") := NULL]

#Summerize Fatalities
Fatalities <- StormData[,sum(FATALITIES, na.rm = TRUE), EVTYPE][,Event := EVTYPE][,Fatalities := V1][,c("EVTYPE","V1") := NULL]

#Summerize Injuries
Injuries <- StormData[,sum(INJURIES, na.rm = TRUE), EVTYPE][,Event := EVTYPE][,Injuries := V1][,c("EVTYPE","V1") := NULL]
```

### 2.5: Merge Data, Clean Enviroment

```{r pressure, echo=FALSE}
#Merge Economic Damage, Fatalities, and Injuries
StormData <- merge(Damage, Fatalities, by = "Event")
StormData <- merge(StormData, Injuries, by = "Event")

#All a label with millions, billions denoted
StormData[ ,Label := sapply(Economic_Damage, milbil)]

#Clean Enviroment
rm(list=setdiff(ls(), "StormData"))
```

3: Plot Charts
------------------

### 3.0: Plot Economic Damage

```{r pressure, echo=FALSE}
# Plot Economic Damage
ggplot(StormData, aes( x = reorder(Event, - Economic_Damage), y= Economic_Damage)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Label),  vjust=-0.25)+
  scale_y_log10(breaks=c(10^0,10^1,10^2,10^3,10^4,10^5,10^6,10^7,10^8,10^9, 10^10, 10^11))+
  labs( title = "Total Economic Damage by Event", x = "Event", y = "Economic Damage")+
  theme(plot.title = element_text(hjust = 0.5))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90))
```

### 3.1: Plot Injuries

```{r pressure, echo=FALSE}
# Plot Injuries
ggplot(StormData, aes( x = reorder(Event, - Injuries), y= Injuries)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Injuries),  vjust=-0.25)+
  scale_y_log10(breaks=c(10^0,10^1,10^2,10^3,10^4,10^5,10^6))+
  labs( title = "Total Injuries by Event", x = "Event", y = "# of Injuries")+
  theme(plot.title = element_text(hjust = 0.5))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90))
```

### 3.1: Plot Fatalities

```{r pressure, echo=FALSE}
# Plot Fatalities
ggplot(StormData, aes( x = reorder(Event, - Fatalities), y= Fatalities)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Fatalities),  vjust=-0.25)+
  scale_y_log10(breaks=c(10^0,10^1,10^2,10^3,10^4))+
  labs( title = "Total Fatalities by Event", x = "Event", y = "# of Fatalities")+
  theme(plot.title = element_text(hjust = 0.5))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90))
```